---
title: "Lipidomics Data Analysis"
subtitle: "iSLS9 Workshop 2021"
author: "Bo Burla and Hyungwon Choi"
date: '2021-02-26'
output:
  html_document:
    keep_md: no
    df_print: paged
  html_notebook: default
  slidy_presentation: default
  ioslides_presentation: default
  word_document: default
bibliography: references.bib
---

# Part 1

# Prerequisites

We will be using tidyverse packages for this part. The `tidyverse` package installs and loads following packages: `ggplot`, `dplyr`, `tidyr`, `readr`, `purrr`, `tibble`, `stringr`, `foracts`. We will also be using `broom` for one module. Furthermore, in one module the package `rgoslin` is used to convert lipid names. This package `rgoslin` is only available via github ((<https://github.com/lifs-tools/rgoslin>). See script below on how to install it. Please note that on Windows, you will need [rtools](https://cran.r-project.org/bin/windows/Rtools/), or on macOS [XCode](https://apps.apple.com/sg/app/xcode/id497799835?mt=12), installed in order to install this package.

We also suggest to install the currently newest version RStudio Version (Version 1.4 or later), which offers a visual RMarkdown/R Notebooks editor.

```{r}
# Run once
#if(!require(devtools)) { install.packages("devtools") }
#install_github("lifs-tools/rgoslin")
```

# Setup

```{r setup}
library(here)
library(tidyverse)
library(broom)
```

# The Lipidomics Dataset

The data used in this workshop is from [Chew *et al*, *JCI Insight,* 2019](https://doi.org/10.1172/jci.insight.126925) The lipdodmics LC-MS raw data used in this publication was reprocessed using MRMkit [Teo *et al*, *Analytical Chemistry*, 2020](https://doi.org/10.1021/acs.analchem.0c03060) [@chew2019; @Teo2020]

## Import and formatting the lipidomics dataset

Read the CSV file and view it. The output is a tibble, which is a data.frame with some additional functions. We see that the first 6 rows contain retention times, m/z values and QC info, which we do not need for now

```{r read csv}
d_orig <- readr::read_csv(here("Part_1/data/quant_table_with_QC.csv"), 
                          col_names = TRUE, trim_ws = TRUE)
d_orig
```

Let's remove these rows

```{r remove annot}
(dat1 <- d_orig %>% slice(-1:-6))
```

The lipid species names still contain detail about the measured transition (-H2O) and the file names contain the file extention .mzML. Let's remove these info from the column names. We also change the column types of `type` to `factor`, and `batch` to `integer`. Furthermore we obtain the `sampleID` from the `filename` of the samples, which has following structure: sampleID\#analysisNo

```{r data cleanup}
d_cleaned <- dat1 %>% 
  rename_with( ~ str_replace(.x, fixed("(-H20)"), "")) %>% 
  mutate(filename = str_replace(
    filename, fixed(".mzML"), ""),
    type = as_factor(type),
    batch = as.integer(batch))

# Extract sampleID using regular expressions and move new column to front
d_cleaned <- d_cleaned %>% 
  mutate(SampleID = str_extract(filename, ".+(?=\\#)")) 

# Assuming that the imported data is in the original analytical sequence, we add an the "run id", which will be used later

d_cleaned <- d_cleaned %>% 
  mutate(run_id = row_number())

# reorder colums, i.e. move the newly generated run_id to the front
d_cleaned <- d_cleaned %>% 
  dplyr::select(filename, run_id, SampleID, everything())
```

## Quality control of the dataset

This dataset contains also 2 QC types: BQCs (Batch QC, pooled plasma co-extracted and measured with samples) and TQCs (technical QC, pooled extract measured in regular intervals). We now would like calculate the %CV (coefficient of variation) of all samples for each lipid species. This be done in the present format of the data, another way, maybe easier approach, is to first convert the table first into a 'long' format:

```{r convert to long, max.print=20, paged.print=TRUE}
# Reshape the table to the long format
d_long <- d_cleaned %>% 
  pivot_longer(names_to = "lipid", values_to = "conc", cols = -filename:-type) %>% 
  arrange(lipid)
d_long  

```

Then we calculate the CVs by first grouping the data by lipid species and QC type and then calculate the CV for each group. Reshaping the table so that the CVs for BQC, TQC split to different columns will make it easier read. Finally, we plot the obtained CVs as a histogram.

```{r calculate CV, max.print=20, paged.print=TRUE}
d_sum <- d_long %>%
  group_by(lipid, type) %>% 
  summarise(CV = sd(conc)/mean(conc) * 100, .groups = "drop")
d_sum

d_sum_w <- d_sum %>% pivot_wider(names_from = type, values_from = CV) 
d_sum_w

summary(d_sum_w)

# We only want to comapre the BQC and Samples, so we exclude TQCs
ggplot(d_sum %>% filter(type != "TQC"), aes(x=CV, color=type)) +
geom_histogram(fill="white", position="dodge", bins = 40)

```

## Plotting run order *vs* concentration

```{r run scatterplot, fig.height=10, fig.width=10}

# We select here the "4" ceramides for plottings
d_plt <- d_long %>% 
  filter(lipid %in% c("Cer d18:1/C16:0", "Cer d18:1/C18:0", "Cer d18:1/C24:0", "Cer d18:1/C24:1"))

# the layer of the QCtype points is defined by the order of the levels in the factor, 
# so lets re-arrange them to move BQC on top and to ensure the scales (color, shapes) are correctly mapped

d_plt$type <- forcats::fct_relevel(d_plt$type, "SAMPLE", "TQC", "BQC")
d_plt <- d_plt %>% arrange(type)

# we will use run_id column that was generated at the beginning
 p <- ggplot(d_plt, aes(x=run_id, y=conc)) + 
        geom_point(aes(colour = type, fill = type, shape  = type),
                   size= 0.4, alpha=0.5, stroke = 1) +
        scale_shape_manual(na.value = NA, values = c(10, 15, 16)) +
        scale_fill_manual(na.value = NA, values= c("#aaaeaf", "#1854f9", "#db0202")) +
        scale_colour_manual(values = c("#aaaeaf", "#1854f9", "#db0202")) +
        facet_wrap(~lipid, ncol = 1, nrow = 4, scales="free_y") +
        aes(ymin=0) +
        xlab("Injection number") +
        theme_light() 
#p  

ggsave(plot = p, 
       filename = here("Part_1/output", "run_scatter_plot.pdf"), 
       units = "mm", 
       width=150, height=200)
```

## Converting lipid species names to the updated LIPID MAPS nomenclature and retrieving lipid annotations

LIPID MAPS has recently updated the lipid species nomenclature to a more comprehensive and standardized system [@liebisch2020]. There is an increasing number of publications adopting this new nomenclature. There is an online service to convert lipid names (<https://apps.lifs.isas.de/goslin/>). The software behind this tool [@Kopczynski2020] is also available in R ([rgoslin](https://github.com/lifs-tools/rgoslin)) and Python ([pygoslin](https://github.com/lifs-tools/pygoslin)).

We will be using `rgoslin` to convert the names. Please see Prerequisites on how to install `rgoslin`. First we retrieve the lipid species names, then we have still need to tweak a bit the current lipid names to a more common format so that goslin can translate them, i.e. remove the C in front of the fatty acid chain info, and convert MHCer and DHCer to Hex1Cer and Hex2Cer, respectively. We will use regular expressions for one case, see also the RStudio cheat sheet. See pre

```{r rgoslin, message=FALSE}
## Installing rgoslin (see also text)
#if(!require(devtools)) { install.packages("devtools") }
#install_github("lifs-tools/rgoslin")
library(rgoslin)
  
# Get a tibble with the lipid names as rows from the original data. Include also retention times and mz values, see later.
# Transposing data frames/tibbles need some tweak.. any better solutions?
d_species <- tibble(species_name = names(d_orig[,-1:-3]), 
                      RT = unlist(d_orig[1,-1:-3]), 
                      Q1 = unlist(d_orig[2,-1:-3]),
                      Q3 = unlist(d_orig[3,-1:-3]))

# Cleaning lipid species names to a more standardized format, that can also be parsed by goslin 
d_species <- d_species %>%  
  mutate(
    species_name_original = species_name,
    species_name = str_replace(species_name, fixed("(-H20)"),""),
    species_name = str_replace(species_name, "\\/C(?=\\d)", "/"),
    species_name = str_replace(species_name, fixed("MHCer"), "HexCer"),
    species_name = str_replace(species_name, fixed("DHCer"), "Hex2Cer"), 
    species_name = str_replace(species_name, fixed("Sphd"), "Sph d"))
  
#d_species

# get new names and a lot more information
d_goslin <- rgoslin::parseLipidNamesWithGrammar(unique(d_species$species_name), "LipidMaps")
#d_goslin

# List species names that could not be parsed by goslin
setdiff(d_species$species_name, d_goslin$`Original Name`)

# Convert to tibble and change column names to syntactically valid names (i.e no spaces) and
# convert selected columns to numeric..
d_goslin_t <- d_goslin %>% 
  as_tibble(.name_repair = "universal") %>% 
  mutate_at(vars(Total.C, Total.OH, Total.DB, Mass), as.numeric)

d_species_annot <- d_species %>%  
  full_join(d_goslin_t, by=c("species_name"="Original.Name"))
d_species_annot
```

## Plotting *Equivalent Carbon Number (ECN)* vs. *Retention Time*

In reversed phase (RP), retention times of lipids increase with the total carbon number and decrease with the total double bond number. The Equivalent Carbon Number (ECN) is an established simple model to prediction of retention times of lipids in RP [@triebl2017a] We now first calculate to ECN based on the total C and DB number obtained from goslin. The formula is $ECN = C-k*DB$, whereby $k$ is 0.5 for the LC method used. Then we plot the ECN against the averaged retention times for each species.

```{r ECNvsRT}
d_species_annot <- d_species_annot %>% 
  mutate(ECN = Total.C - 0.5 * Total.DB)

ggplot(d_species_annot, aes(x = ECN,  y = RT, color = Functional.Class.Abbr)) +
  geom_point(size= 2, alpha = 0.9) +
  scale_color_brewer(palette="Dark2") + 
   geom_smooth(method = "lm", formula = y ~ x, se = FALSE)
```

## Performing a simple t test on all lipid species between two groups of samples

In this example we want to test for lipid species that have different plasma concentration between between males and females. We will exclude subject that developed T2DM in this cohort.

#### Prepare the data and get p-values and log2FC for each lipid species (method 1)

```{r getFC_P}

# Load table with sample annotations
d_annot <- readr::read_csv(here("Part_1/data/sample_annotations.csv"), col_names = TRUE, col_types = cols(.default = col_number()), trim_ws = TRUE)

d_annot$SampleID <- as.character(d_annot$SampleID)

# Select only subjects that had no T2D and re-code gender from 1/2 to M/F 
d_annot_sel <- d_annot %>% 
  filter(DM_incidence == 0) %>%
  dplyr::select(SampleID, Gender) %>%
  mutate(Gender = ifelse(Gender == 1, "M", "F"))

# Add Gender to each data point (=subject/lipid) of the table. Right join means only SampleID is the 2nd (=right) table are selected 
d_stat <- d_long %>% 
  right_join(d_annot_sel, by = "SampleID")

# Check for NAs in species/samples with NAs and drop these entries
d_stat %>%
  dplyr::select(SampleID, lipid, conc, Gender) %>% 
  summarise_all(funs(sum(is.na(.))))

d_stat <- d_stat %>% drop_na()

# Calculate log2FC and p values from t.test contrasting M v F
# METHOD 1 using 'classic' dplyr group_by and summarise 
d_stat_sum <- d_stat %>% 
  group_by(lipid) %>% 
  summarise(
    log2FC = log2(mean(conc[Gender == "F"])/mean(conc[Gender == "M"])),
    p_value = t.test(x = conc[Gender == "F"], 
                     y = conc[Gender == "M"],
                     paired = FALSE,
                     var.equal = FALSE)$p.value)              

# Multiple testing correction (FDR)
d_stat_sum$FDR <- p.adjust(d_stat_sum$p_value)


hist(d_stat_sum$p_value, breaks = 30)
```

#### Using `purrr` and `broom` to get comprehensive t.test results for all species (method 2)

In the above script we only retrieved the *p* value from the *t* test. If we want to retrieve additional parameter such as *t* and *stderr,* we would need to run the t test function several times for each species or write a function to extract these information. The package broom offers functions to convert the results of various tests, including `t.test`, `lm`*,* `cor.test` and `lmer` to a flat tables, making it easier to be used. Together with functions from purrr, this quite useful when performing repeated tests, as for example running the same test over hundreds of different lipid species.

Let us first have a look at the output of a `t.test`:

```{r purrr}
res <- t.test(x = c(11,21,13,5,15), c(33,46,39,41,51))
res
```

Now we use `purrr`'s `tidy` function to parse the `t.test` results into a flat table:

```{r tidy}
broom::tidy(res)
```

To run a test over all lipid species of our dataset we first group our datasets, each group containing the data of one species. We obtain data.frame/tibble with column 'data'.

```{r nest}
# First we filter the datasets for the levels we want to compared (M vs F)
d_filt <- d_stat %>% filter(Gender %in% c("M", "F"))
    
# Group the dataset into lipid species subsets  
d_nest <- d_filt %>% 
  group_nest(lipid) 

d_nest

```

Each cell in 'data' column contains a table with the data for one species

```{r shownested}
d_nest$data[3]
```

Using the `purrr` map function we can walk through all the sub-tables (one for each species) and run t.test and covert it to a tidy result. The results are in the column 'ttest', in the format of 1-row tables (from the output of `tidy`)

```{r mapttest}
d_ttest <- d_nest %>%
    mutate(
      ttest = purrr::map(data, ~ tidy(t.test(conc ~ Gender, data = .x)))
      )
d_ttest
```

We can now "flatten" those 1 row tables using the purrr

```{r unnest}
d_ttest %>% unnest(ttest)
```

Here again in one block, including calculations of log2(FC) values. There are many variations to code this using purrr, glad to learn other approaches.

```{r ttestall}

d_ttest2 <- d_stat %>% 
  filter(Gender %in% c("M", "F")) %>% 
  group_nest(lipid) %>% 
  mutate(
    ttest = purrr::map(data, ~ tidy(t.test(conc ~ Gender, data = .x))),
    log2FC = purrr::map_dbl(data, ~ log2(mean(.x$conc[.x$Gender == "F"]) / 
                                         mean(.x$conc[.x$Gender == "M"])))) %>%
  unnest(ttest)

# Multiple testing correction (FDR)
d_ttest2$FDR <- p.adjust(d_ttest2$p.value)

d_ttest2
```

## Plot *P* values and FC as a Volcano plot

To end this part of the workshop, we will create a volcano plot using the data from last block.

```{r volcano}

library(ggrepel)  # used to 'smartly' move the labels

# Set tresholds for FC and p value 
FC_MIN <- 1.2
FDR_MAX <- 0.01

# add columns to the data.frame definin 'significantly different' species
d_stat_sum <- d_ttest2 %>% 
  mutate(
    sign = case_when(
      log2FC > log2(FC_MIN) & FDR < FDR_MAX ~ "Higher in F",
      log2FC < -log2(FC_MIN) & FDR < FDR_MAX ~ "Lower in F",
      TRUE ~ "n.s."),
    label_sign =  ifelse(sign != "n.s.", lipid, "")
  )



ggplot(data=d_stat_sum, aes(x= log2FC, y= -log10(FDR), color = sign, label = label_sign)) +
  geom_point() + 
  geom_text_repel() +
  scale_color_manual(values=c("red", "blue", "black")) +
  geom_vline(xintercept=c(0, -log2(FC_MIN), log2(FC_MIN)), 
             col="black", linetype = "dotted", size = 1) +
  geom_hline(yintercept=-log10(FDR_MAX), col="red") +
    theme_minimal()
```

# References
